\section{Solving Equations} \label{sec:lin_eqn}

The advent of linear algebra was mainly to solve coupled equations with more than one variable. Even 4000 years ago, people were fascinated by solving 2x2 matrix equations that represented two equations in 2 variables. After that, small progress was made, and solving 3x3 matrices was introduced by the Chinese. But not until the late 17th century were any advancements in linear algebra mathematics. Leibneitz first started to formalize and advance this field. Gauss introduced the Gaussian elimination method, a generalized and easy method to handle equations in multiple variables. Later, Cayley named the arrays of numbers matrix because they were seen to be the generators of determinants. His works introduced the notion of the identity matrix and also the notion of the inverse of matrices. Linear algebra gradually became more popular as computers were developed, and solving differential equations was done approximately using linear algebra. 

\subsection{The advent of Computers}

The advent of computers and computer science revolutionized the field of mathematics and took linear algebra to new levels. We use linear algebra for many things today, including solving mundane linear equations and differential equations. Solving differential equations opens the doors for the analysis of complex systems and time analysis. It opens the door for varied uses, such as in physics, quantum mechanics, market analysis, and machine learning. Most of the advanced machine learning models and algorithms that are in use today have been developed from linear algebra. So, even artificial intelligence is using linear algebra. Linear algebra has eventually become a language of computers. 

\subsection{From a student's point of view}

The first use of primitive linear algebra that we do is in the word problems that give rise to coupled linear equations such as:

\begin{equation}
    5x+3y=6; 2x+7y=9
\end{equation}

this can be represented as two equations in two variables but also as a matrix equation, more precisely:

\begin{equation}
    \begin{pmatrix}
        5 & 3\\
        2 & 7
    \end{pmatrix}\begin{pmatrix}
        x\\y
    \end{pmatrix}=\begin{pmatrix}
        6 \\9
    \end{pmatrix}
\end{equation}

it can be written as the matrix equation $Ax=b$

so this problem is simplified to this generalized matrix equation. The whole of linear algebra is about talking about the matrix equation: $Ax=b$ 